\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-24 NAR_DRAFT.tex
comment something
More minor revisions and comments.

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}\textbf{A}-\textbf{D}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 without the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
\begin{equation}
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
\end{equation}
Returning to Eq. \ref{eqn:detailed_prior}, given that a mutation has occurred, the posterior probability it occurred in context $c$ is
\begin{equation}
  \label{eqn:post_pred}
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$ is then
\begin{equation}
  \label{eqn:to_allele}
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
\end{equation}
The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
The R implementation of \batcave ships with pre-computed tables for both whole exomes and whole genomes.

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We can then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\citep{Williams2016}.
%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

@@

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package (batcaver).
The package leverages the Bioconductor packages \textrm{BSgenome}, \textrm{GenomicAlignments}, \textrm{VariantAnnotation}, and \textrm{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification. 
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

@@

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp 
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
\rngcomment{Did MuTect fail to call anything we know is there?}
\bkmcomment{Depends on the cutoff. This is the precision-recall curve.}

\subsection{Calibration metric}
The quantify the difference in calibration between MuTect and \batcave, we use the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1,False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$, and the Integrated Calibration Index is the average distance between the fit and that line.

\section{RESULTS}
We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites in to high- and low-confidence groups.
The mutation profile is estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500x exome and 2,000 seconds for a 100x whole genome.

To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}\textbf{A-C}).
We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{Convergence of the site-specific prior to the data generating distribution. Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
  }
\label{NAR-kl_fig}
\end{figure}

\subsection{Convergence of the prior}
To improve performance, the context-dependent prior probability of mutation must converge to a good representation of the data generating distribution within the set of high-confidence mutations.
When applied to simulated whole exome and whole genome data, the prior converges within a few hundred mutations (Fig.~\ref{NAR-kl_fig}).
For comparison, in our simulated data sets the number of high-confidence mutations ranges between 1500 and 5000, and for a real acute myeloid leukemia it is over 17,000~\cite{Griffith2015}.

>>>>

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.

To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 without the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
\begin{equation}
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\},
\end{equation}
and $I$ is the indicator function.
Returning to Eq. \ref{eqn:detailed_prior}, given that a mutation has occurred, the posterior probability it occurred in context $c$ is
\begin{equation}
  \label{eqn:post_pred}
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$ is then
\begin{equation}
  \label{eqn:to_allele}
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
\end{equation}

The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
The R implementation of \batcave ships with pre-computed tables for both human whole exomes and whole genomes.

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\citep{Williams2016}.
%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

@@

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package (batcaver).
The package leverages the Bioconductor packages \textrm{BSgenome}, \textrm{GenomicAlignments}, \textrm{VariantAnnotation}, and \textrm{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.  \rngcomment{Add citations to all tools that have them.}

@@

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of alleles.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp 
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
\rngcomment{What are we doing about the multi-regional nature of this data? Are we combining across regions somehow?}
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

\subsection{Calibration metric}
To quantify the difference in calibration between MuTect and \batcave, we use the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
The Integrated Calibration Index is the average distance between the fit and that line.
\rngcomment{This can't be true, otherwise perfect calibration would be ICI=1... What is it?}

\section{RESULTS}
We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
The mutation profile is estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.

To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100X whole genomes and 500X whole exomes for three different mutation profiles.
The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NAR-simsigfig}\textbf{A-C}).
We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
Because our focus is on evaluating the statistical calling model, we compute all metrics using only those potential variants that pass MuTect's heuristic filters.

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{Convergence of the site-specific prior to the data generating distribution. Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations for whole exomes. Convergence for whole genomes is similar.
  \rngcomment{Omit panel B and remove A label. These are similar enough we don't need to show both.}
  }
\label{NAR-kl_fig}
\end{figure}

% XXX: RNG stopped here...

\subsection{Convergence of the prior}
To improve performance, the context-dependent prior probability of mutation must converge to a good representation of the data generating distribution within the set of high-confidence mutations.
When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}).
For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.
